---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Arrange, Group By, Summarize

## Load packages

```{r}
_____(tidyverse)
```

## Import data into R and read into data frame

It may take a few moments for this file to load.

```{r}
#Read in the csv.
sqf_url <- "https://www1.nyc.gov/assets/nypd/downloads/zip/analysis_and_planning/stop-question-frisk/sqf-2011-csv.zip"
temp <- tempfile()
download.file(sqf_url, temp)
sqf_zip <- unzip(temp, "2011.csv")
sqf_2011 <- read.csv(sqf_zip, stringsAsFactors = FALSE) 
sqf_2011_race_cat <- read.csv("https://raw.githubusercontent.com/lindsaypoirier/STS-101/master/Data/SQF/sqf_race_categories.csv", stringsAsFactors = FALSE) 
```

```{r}
sqf_2011 <- sqf_2011 %>% left_join(sqf_2011_race_cat, by = "race")
```

Added variables for 1) weapon found, 2) arrestsumm, 3) forceused

```{r}
sqf_2011 <- 
  sqf_2011 %>%
  mutate(wpnfound=ifelse((pistol=="Y"|riflshot=="Y"|asltweap=="Y"|knifcuti=="Y"|machgun=="Y"|othrweap=="Y"),"Y","N"))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(arrestsumm=ifelse((sumissue=="Y"|arstmade=="Y"),"Y","N"))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(forceused=ifelse((pf_hands=="Y"|pf_wall=="Y"|pf_grnd=="Y"|pf_drwep=="Y"|pf_ptwep=="Y"|pf_baton=="Y"|pf_hcuff=="Y"|pf_pepsp=="Y"|pf_other=="Y"),"Y","N"))

```

### Summarize

What was the age of the oldest individual stopped? What was the age of the youngest individual stopped?

In 2011, how many stops in which an individual was frisked resulted in a weapon found? What percentage of the frisk stops resulted in a weapon found?

How many stops occurred in 2011?

In 2011, how many stops involved a frisk? 

In 2011, for how many stops was the officer uniformed? 

In 2011, how many stops resulted in an arrest or summons? What percentage of the total stops resulted in an arrest or summons?

In 2011, how many stopped individuals refused to present identification?

## Group by

How many individuals of each race were stopped in 2011?

How many individuals of each sex were stopped in 2011?

How many individuals of each sex were stopped in instances where force was used?

For each racial group represented in the dataset, what were the 1) total number of frisks, 2) total numbers of stops, and 3) percentage of stops that resulted in a frisk.

For each racial group represented in the dataset, what were the 1) total number of instances a weapon was found, 2) total numbers of stops, and 3) percentage of stops that resulted in a weapon found.

## Arrange

How can I order the plots above so that releases decrease from most to least?

## Top n

Which police precinct had the greatest number of stops? Was this the same precinct with the greatest number of frisks?

## How do we plot this?

Plot NYPD Stops by Age in 2011

Plot NYPD Stops by Age 100+ in 2011

Plot Percent Stops Resulting in Frisk by Race in 2011

Plot Stops Resulting in Frisk by Race in 2011

Plot Percent Frisks Resulting in Weapon Found by Race in 2011

Plot Frisks Resulting in Weapon Found by Race in 2011

