---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

# Arrange, Group By, Summarize

## Load packages

```{r}
_____(tidyverse)
```

## Import data into R and read into data frame

It may take a few moments for this file to load.

```{r}
#Read in the csv.
tri_2019 <- _____("https://raw.githubusercontent.com/lindsaypoirier/STS-101/master/Data/TRI/TRI_2019_US.csv")
```

## Review

How can I filter to California TRI facilities?

How can I plot the total releases in California TRI facilities by county?

How can I plot the total releases in California TRI facilities by county, divided by industry?

### Summarize

How can I calculate the total releases across all facilities and chemicals?

How can I calculate the total releases of Benzene?

How can I calculate the total releases in zip code 95616 of any chemical?

How can I calculate the total releases of Lead in California?

How can I calculate the total releases of any chemical by the Petroleum industry?

## Group by

How can I calculate the total releases per industry?

How can I calculate the total releases per chemical?

## Arrange

How can I order the plots above so that releases descrease from most to least?

## Top n

How can I find the top three most polluting facilities in the dataset?
