---
title: "Categories and Filtering"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Categories and Filtering

## Load packages

```{r}
_____(tidyverse)
```

## Import data into R and read into data frame

It may take a few moments for this file to load.

```{r}
#Read in the csv.
tri_2019 <- _____("https://raw.githubusercontent.com/lindsaypoirier/STS-101/master/Data/TRI/TRI_2019_US.csv")
```

## Review

How can I see the first six rows of this data?

How can I check the dimensions of this data?

How can I check the column names of this data?

What does each observation (row) in this dataset refer to? : 

How can I check the distinct chemicals in this dataset?

How can I check the distinct facilities in this dataset?

How can I check the distinct industry codes in this dataset?

## New Material

### Count

What if I wanted to know how many times each chemical appeared in the dataset? What does this number tell me?

What if I wanted to know how many times each facility appeared in the dataset? What does this number tell me?

### Filter

How can I find all of the facilities in zip code 95616?

How can I find all of the facilities that have released more than 1000 pounds of any chemical?

How can I find all of the facilities with NA values in Total Releases?

How can I find the facility with the max total releases of any chemical?

### Summarize

How can I calculate the total releases across all facilities and chemicals?

How can I calculate the total releases of Benzene?

How can I calculate the total releases in zip code 95616 of any chemical?

How can I calculate the total releases of Lead in Califoria?

How can I calculate the total releases of any chemical by the Petroleum industry?

### ggplot

How can I plot the total releases by carcinogen vs. non-carcinogen?

> Be sure to add a title, labels, and a caption.

How can I plot the total releases by industry?

> Note: Will need + coord_flip() and + scale_y_continuous(labels = scales::comma)

How can I plot the total releases by industry code?

How can I divide that plot by carcinogen and non-carcinogen chemicals?

## Data Exploration

