---
title: "Bridge Conditions in California and New York"
author: Professor Lindsay Poirier
output: html_document
---


```{r}
library(tidyverse)
```


```{r}

nbi_ca<-read.csv("NBI/nbi_CA18.csv")
nbi_ny<-read.csv("NBI/nbi_NY18.csv", quote = "")
nbi_fips_lookup<-read.csv("NBI/fips.csv")
nbi_mat_lookup<-read.csv("NBI/nbi_mat_lookup.csv")
nbi_ser_lookup<-read.csv("NBI/nbi_service_lookup.csv")
nbi_age_lookup<-read.csv("NBI/nbi_agency_lookup.csv")

#Added categorical labels associated with numeric labels
nbi_ca <- nbi_ca %>% left_join(nbi_mat_lookup)
nbi_ny <- nbi_ny %>% left_join(nbi_mat_lookup)

nbi_ca <- nbi_ca %>% left_join(nbi_ser_lookup)
nbi_ny <- nbi_ny %>% left_join(nbi_ser_lookup)

nbi_ca <- nbi_ca %>% left_join(nbi_age_lookup)
nbi_ny <- nbi_ny %>% left_join(nbi_age_lookup)

#Formatted County Code field to match FIPS and joined county names
nbi_ca$COUNTY_CODE_003<-str_pad(nbi_ca$COUNTY_CODE_003, 3, pad = "0")
nbi_ny$COUNTY_CODE_003<-str_pad(nbi_ny$COUNTY_CODE_003, 3, pad = "0")
nbi_fips_lookup$FIPS<-as.character(nbi_fips_lookup$FIPS)
nbi_ca <- nbi_ca %>% mutate(FIPS = paste(STATE_CODE_001, COUNTY_CODE_003, sep="")) %>% left_join(nbi_fips_lookup) 
nbi_ny <- nbi_ny %>% mutate(FIPS = paste(STATE_CODE_001, COUNTY_CODE_003, sep="")) %>% left_join(nbi_fips_lookup)

```

```{r}
#Changed all instances of "N" to NA
is.na(nbi_ca) <- nbi_ca == "N"
is.na(nbi_ca) <- nbi_ca == "N "
is.na(nbi_ny) <- nbi_ny == "N"
is.na(nbi_ny) <- nbi_ny == "N "
```

```{r}
#Converted character columns to numeric
cols<-c(32, 37:40, 42, 67:71, 76:78, 80)
nbi_ca[,cols]<-sapply(nbi_ca[,cols], function(x) as.numeric(x))
nbi_ny[,cols]<-sapply(nbi_ny[,cols], function(x) as.numeric(x))
```

```{r}
nbi_ca$TOTAL_IMP_COST_096 <- nbi_ca$TOTAL_IMP_COST_096*1000
nbi_ny$TOTAL_IMP_COST_096 <- nbi_ny$TOTAL_IMP_COST_096*1000
```

Today, we will be working with a series of charts that are useful in exploratory data analysis (EDA). EDA is a creative practice where we get to know data by examining the variation between variables. Variation is a term used to describe how much measurements change from one observation (or row) to another. One way we can examine variation is by looking at the distributions of values across variables (and eventually multiple variables) - or in other words how many times each value is present within a column in our dataset. (Remember we have being producing tables like this with group_by(VARIABLE) %>% summarize(total = n())). To analyze distributions, we will work with the following plot types today. 
1. geom_bar()
2. geom_histogram()
3. geom_freqpoly()
4. geom_boxplot()

geom_bar() creates a bar chart that displays the number of times a value is present within a given categorical variable. When using geom_bar(), we supply a categorical variable in our dataset as an x-axis aeshetic and ggplot creates bins (or groups) of all the common values in that column and plots them on the x-axis. ggplot then plots bars for each value - with the height of the bar corresponding with the number of times a value appears in that column of the dataset. When calling geom_bar(), we need only supply one variable name as an aesthetic; the y-axis becomes the count.

How many bridges were in each county in CA and NY 2018?
```{r}
nbi_ca %>%
  ____(_____(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 CA Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))


nbi_ny %>%
  _____(_____(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 NY Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```

By faceting the data, or by adding fill or color properties, we can begin to show variation in the data across more than one variable. 

For each bridge material type, how many bridges in CA and NY received each structural evaluation?
```{r}
nbi_ca %>%
  _____(_____(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 CA Bridges by Structural Evaluation and Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  _____(~_____)

nbi_ny %>%
  _____(_____(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 NY Bridges by Structural Evaluation and Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  _____(~_____)
```

For each party responsible for bridge maintenance, how many bridges in CA and NY received each structural evaluation?
```{r}
nbi_ca %>%
  _____(aes(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 CA Bridges by Responsible Agency") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) 

nbi_ny %>%
  _____(aes(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 NY Bridges by Responsible Agency") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) 


nbi_ca %>%
  _____(aes(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 CA Bridges by Responsible Agency") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  _____(~RESPONSIBLE_AGENCY)

nbi_ny %>%
  _____(aes(x = _____)) + 
  _____ + 
  labs(title = "Number of 2018 NY Bridges by Responsible Agency") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  _____(~RESPONSIBLE_AGENCY)
```

When we use geom_bar(), the width of each bar will always be equal to 1. This is appropriate when we are plotting counts of categorical variables since, each variable will be distinct. However sometimes, we want to count how many of value are present in a continuous variable. In such cases, we can use geom_histogram(). geom_histogram() is very similar to geom_bar() except in that it allows you to set a binwidth. Binning data refers to the creation of bins in which to group numerical values falling into a certain range, and the binwidth designates the length of that range. Let's say I have the values 0.1, 0.2, 0.5, 1.2, 3.5, 3.6. When I set a binwidth to 1, it will create the following bins:
0 (including 0.1, 0.2, 0.5)
1 (including 1.2)
2 (empty)
3 (including 3.5, 3.6)

If I set the binwidth to 0.5, it will create the following bins:

0 (including 0.1, 0.2)
0.5 (including 0.5)
1 (including 1.2)
1.5 (empty)
2 (empty)
2.5 (empty)
3 (empty)
3.5 (including 3.5, 3.6)

Setting the binwidth to a lower number allows us to view the data in more fine-grained detail, while setting it to a higher number smooths out some anamolies to show trends. 

How many bridges were built each year in CA and NY?
```{r}
nbi_ca %>%
  ggplot(aes(x = _____)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Year Built")

nbi_ny %>%
  ggplot(aes(x = _____)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Year Built")

nbi_ca %>%
  ggplot(aes(x = YEAR_BUILT_027)) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 CA Bridges by Year Built")

nbi_ny %>%
  ggplot(aes(x = YEAR_BUILT_027)) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 NY Bridges by Year Built")
```

How many bridges received each structural evaluation in CA and NY?
```{r}
nbi_ca %>%
  _____(aes(x = _____)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Structural Evaluation")

nbi_ny %>%
  _____(aes(x = _____)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Structural Evaluation")
```

geom_bar() and geom_histogram() displays distributions as bar charts. We can also view distrubtions as line charts, which is often useful when we want to view the distributions across an additional categorical variable. To do this, we will use geom_freqpoly(). With freqpoly, the height of the line corresponds with the number of times a range of values within a spefic bin appears in the supplied x-aeshetic variable (or column). 

How many bridges of each material type were built each year in CA and NY?
```{r}
nbi_ca %>%
  ggplot(aes(x = _____, col = _____) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Year Built and Structure Kind")

nbi_ny %>%
  ggplot(aes(x = _____, col = _____)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Year Built and Structure Kind")

nbi_ca %>%
  ggplot(aes(x = _____, col = _____) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 CA Bridges by Year Built and Structure Kind")

nbi_ny %>%
  ggplot(aes(x = _____, col = _____)) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 NY Bridges by Year Built and Structure Kind")
```

For each strucutral evaluation, how many bridges were built each year in CA and NY?
```{r}
nbi_ca %>%
  ggplot(aes(x = YEAR_BUILT_027, col = as.factor(STRUCTURAL_EVAL_067))) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Year Built and Structural Evaluation")

nbi_ny %>%
  ggplot(aes(x = YEAR_BUILT_027, col = as.factor(STRUCTURAL_EVAL_067))) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Year Built and Structural Evaluation")

nbi_ca %>%
  ggplot(aes(x = YEAR_BUILT_027, col = as.factor(STRUCTURAL_EVAL_067))) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 CA Bridges by Year Built and Structural Evaluation")

nbi_ny %>%
  ggplot(aes(x = YEAR_BUILT_027, col = as.factor(STRUCTURAL_EVAL_067))) + 
  _____(binwidth = 5) + 
  labs(title = "Number of 2018 NY Bridges by Year Built and Structural Evaluation")
```

For each service type, how many bridges received in structural evaluation in CA and NY?
```{r}
nbi_ca %>%
  ggplot(aes(x = STRUCTURAL_EVAL_067, col = SERVICE)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Structural Evaluation and Service Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))

nbi_ny %>%
  ggplot(aes(x = STRUCTURAL_EVAL_067, col = SERVICE)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Structural Evaluation and Service Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) 


#Here we change the scale of the y axis to logarithmic
nbi_ca %>%
  ggplot(aes(x = STRUCTURAL_EVAL_067, col = SERVICE)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 CA Bridges by Structural Evaluation and Service Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_log10()

nbi_ny %>%
  ggplot(aes(x = STRUCTURAL_EVAL_067, col = SERVICE)) + 
  _____(binwidth = 1) + 
  labs(title = "Number of 2018 NY Bridges by Structural Evaluation and Service Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_log10()
```

For each material type, how many bridges of each lengh are there in CA and NY?
```{r}
#Bins have constant width on the transformed scale.
nbi_ca %>%
  ggplot(aes(x = STRUCTURE_LEN_MT_049, col = TYPE)) + 
  geom_freqpoly(binwidth = 0.05) + 
  labs(title = "Number of 2018 CA Bridges by Length and Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_x_log10()

nbi_ny %>%
  ggplot(aes(x = STRUCTURE_LEN_MT_049, col = TYPE)) + 
  geom_freqpoly(binwidth = 0.05) + 
  labs(title = "Number of 2018 NY Bridges by Length and Type") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_x_log10()
```

Which bridges have improvement costs? How many bridges with each sufficiency rating have improvemnt costs?
```{r}
nbi_ca_imp <- 
  nbi_ca %>% 
  mutate(HAS_IMPROVEMENT_COSTS = !is.na(TOTAL_IMP_COST_096)) 

nbi_ny_imp <- 
  nbi_ny %>%
  mutate(HAS_IMPROVEMENT_COSTS = !is.na(TOTAL_IMP_COST_096)) 

nbi_ca_imp %>%
  ggplot(aes(x = SUFFICIENCY_RATING, fill = HAS_IMPROVEMENT_COSTS)) + 
  geom_histogram(binwidth = 5) + 
  labs(title = "Number of 2018 CA Bridges by Sufficiency Rating")

nbi_ny_imp %>%
  ggplot(aes(x = SUFFICIENCY_RATING, fill = HAS_IMPROVEMENT_COSTS)) + 
  geom_histogram(binwidth = 5) + 
  labs(title = "Number of 2018 NY Bridges by Sufficiency Rating")
```

Aside: Check out how much these bridges cost!
```{r}
nbi_ca %>%
  ggplot(aes(x = COUNTY_NAME, y = TOTAL_IMP_COST_096)) +
  _____ + 
  labs(title = "Total Cost of 2018 CA Bridge Improvements by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)

nbi_ny %>%
  ggplot(aes(x = COUNTY_NAME, y = TOTAL_IMP_COST_096)) +
  _____ + 
  labs(title = "Total Cost of 2018 CA Bridge Improvements by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)
```

We can get more detailed information on the distribution of values in a variable by using geom_boxplot(). A boxplot summarizes the distribution of values across a variable into five numbers - the median, the maximum, th minimum, the first quartile, and the third quartile. When calling geom_boxplot(), we provide a catgorical x-aesthetic and a continuous y-aesthetic, allowing us to measure the distribution of values in the y-aesthetic for each categorical value in the x-aesthetic. 

For each county in NY and CA, what is the distribution of sufficiency ratings for bridges?
```{r}
nbi_ca %>%
  ggplot(aes(x = _____, y = _____)) +
  _____ + 
  labs(title = "Sufficiency Ratings of 2018 CA Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))

nbi_ny %>%
  ggplot(aes(x = _____, y = _____)) +
  _____ + 
  labs(title = "Sufficiency Ratings of 2018 NY Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1))
```

For each county in NY and CA, what is the distribution of sufficiency ratings for annual daily traffic of bridges?
```{r}
nbi_ca %>%
  ggplot(aes(x = COUNTY_NAME, y = ADT_029)) +
  _____ + 
  labs(title = "Average Daily Traffic of 2018 CA Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)

nbi_ny %>%
  ggplot(aes(x = COUNTY_NAME, y = ADT_029)) +
  _____ + 
  labs(title = "Average Daily Traffic of 2018 NY Bridges by County") + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)
```


```{r}
nbi_ca %>%
  ggplot(aes(x = RESPONSIBLE_AGENCY, y = TOTAL_IMP_COST_096)) +
  _____ + 
  labs(title = "Average Daily Traffic of 2018 CA Bridges by County") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)

nbi_ny %>%
  ggplot(aes(x = RESPONSIBLE_AGENCY, y = TOTAL_IMP_COST_096)) +
  _____ + 
  labs(title = "Average Daily Traffic of 2018 NY Bridges by County") + 
  theme(axis.text.x = element_text(angle = 45, hjust=1)) + 
  scale_y_continuous(labels = scales::comma)
```





