---
title: "R Review - Stop, Question, and Frisk"
output: html_notebook
---
```{r}
library(tidyverse)
sqf_zip <- unzip("SQF/sqf-2011-csv.zip")
sqf_2011 <- read.csv(sqf_zip) 
```

```{r}
for(i in c(13:15, 17, 21:73, 76:78)){
  sqf_2011[,i]<-ifelse(sqf_2011[,i]=="Y", 1, 0)
}
```

Added variables for 1) weapon found, 2) arrestsumm, 3) forceused
```{r}
sqf_2011 <- 
  sqf_2011 %>%
  mutate(wpnfound=ifelse((pistol==1|riflshot==1|asltweap==1|knifcuti==1|machgun==1|othrweap==1),1,0))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(arrestsumm=ifelse((sumissue==1|arstmade==1),1,0))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(forceused=ifelse((pf_hands==1|pf_wall==1|pf_grnd==1|pf_drwep==1|pf_ptwep==1|pf_baton==1|pf_hcuff==1|pf_pepsp==1|pf_other==1),1,0))

```

How many stops occurred in 2011?
```{r}
sqf_2011 %>%
  _____
```

What was the age of the oldest individual stopped? What was the age of the youngest individual stopped?
```{r}
sqf_2011 %>%
  _____(max_age = _____(_____))

sqf_2011 %>%
  _____

```

In 2011, for how many stops was the officer uniformed? In what percentage of the total stops was the officer uniformed?
```{r}
sqf_2011 %>%
  _____(uniformed = _____(_____))

sqf_2011 %>%
  _____(uniformed = _____(_____)/_____) * 100
```

In 2011, how many stops involved a frisk? What percentage of the total stops involved a frisk?
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____
```

In 2011, how many stops in which an individual was frisked resulted in a weapon found? What percentage of the frisk stops resulted in a weapon found?
```{r}
sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(weapon = _____(_____))

sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(weapon_pct = _____(_____)/_____) * 100
```

In 2011, how many stops resulted in an arrest or summons? What percentage of the total stops resulted in an arrest or summons?
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____
```

In 2011, how many stopped invdividuals refused to present identification?
```{r}
sqf_2011 %>%
  _____(______ == _____) %>%
  _____(refused = _____)
```

How many individuals of each race were stopped in 2011?
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(stopsrace = _____) 
```

How many individuals of each sex were stopped in 2011?
```{r}
sqf_2011 %>%
  _____
```

How many individuals of each sex were stopped in instances where force was used?
```{r}
sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(_____) %>%
  _____(_____ = _____)
```

For each racial group represented in the dataset, what were the 1) total number of frisks, 2) total numbers of stops, and 3) percentage of stops that resulted in a frisk.

For each racial group represented in the dataset, what were the 1) total number of instances a weapon was found, 2) total numbers of stops, and 3) percentage of stops that resulted in a weapon found.
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(frisked=_____(_____), stopped = _____, pctfrisked = _____/_____*100)

sqf_2011 %>%
  _____
```

Which police precinct had the greatest number of stops? Was this the same precinct with the greatest number of frisks?
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(stopspct = _____) %>%
  _____ %>%
  _____(_____ == _____(_____))

sqf_2011 %>%
  _____


```


For each precinct represented in the dataset, what were the 1) total number of frisks, 2) total numbers of stops, and 3) percentage of stops that resulted in a frisk.

For each precinct represented in the dataset, what were the 1) total number of instances a weapon was found, 2) total numbers of stops, and 3) percentage of stops that resulted in a weapon found.
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____

```



