---
title: "R Homework 3"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

Run the code chunk below to load the tidyverse library.
```{r}
library(tidyverse)
library(scales)
```


Run the code chunk below to create the scorecard dataframe. Don't worry about the the "NAs introduced by coercion" listed below. 
```{r}
#Read in the csv.
scorecard <- read.csv("Scorecard/MERGED2017_18_PP.csv", stringsAsFactors = FALSE)

#Convert all values entered as "NULL" to NAs. 
is.na(scorecard) <- scorecard == "NULL"

#Convert columns 12 - 1899 to numeric.
for(i in c(12:1899)){
  scorecard[,i]<-as.numeric(scorecard[,i])
}
```

Part I. Practice Problems 

1. In which 4-year institutions with an undergraduate enrollment of over 300 did more than 95% of full-time first-time degree-seeking undergraduate students complete their degree within 150% of normal time? Display the institution name, the 150% completion rate, the number of full-time first-time degree-seeking undergraduates in the corresponding completion year cohort, and the total undergraduate enrollment.

```{r}
scorecard %>% 
  _____(_____ > _____ & _____ == _____ & _____ > _____) %>%
  _____(_____, _____, _____, _____)
```

2. What are the names, cities, states, and undergraduate enrollments of institutions in "college towns" - those universities with an undergraduate enrollment of over 20,000 that are also in a city or suburb with a population of less than 100,000? How many college towns are in each state?
```{r}
#List the names, cities, states, and undergraduate enrollments of college towns
scorecard %>% 
  filter((_____ == _____ | _____ == _____) & (_____ > _____)) %>% 
  _____(_____, _____, _____, _____) #List

#Calculate the number of college towns in each state
scorecard %>% 
  _____ %>% #copy filter statement above
  _____(_____) %>% 
  _____(total = _____)
```
3. What are the names, cities, states, and undergraduate enrollments of the institution with the largest undergraduate enrollment in each region? What are the names, cities, states, and admission rates for the institution wtih the lowest admission rate in each state?

```{r}
#List the names, cities, states, and undergraduate enrollments of the institution with the largest undergraduate enrollment in each region
scorecard %>% 
  _____(_____) %>%
  _____(_____ == _____(_____, na.rm = TRUE)) %>%
  ungroup() %>%
  _____(_____, _____, _____, _____)

#Following same logic as above list the names, cities, states, and admission rates for the institution wtih the lowest admission rate in each state
scorecard %>%
  _____ %>%
  _____ %>% 
  ungroup() %>%
  _____
```

4. How many public institutions are there in each state? Which state has the greatest number of public institutions? 
```{r}
#Calculate how many public institutions are in each state
scorecard %>%
  _____(______ == _____) %>% 
  _____(_____) %>%
  _____(publicschools = _____)

#Add two lines to the code above to filter the resulte to the state with the greatest number of public institutions
scorecard %>%
  _____ %>% #copy from above
  _____ %>% #copy from above
  _____ %>% #copy from above
  _____ %>%
  _____(_____ == _____(_____))
```
 Part II. Metric Fixation
 
In part two of this assignment, you will design a metric for designating the "best universities for _____ ". For instance, your metric may designate the "best universities for undergraduate women", or the "best universities for Psychology majors". As an example below, I will use the metric: "best universities for Social Science Professors under 30"


1. Select five variables in the scorecard dataset that you believe should be used to assess the quality of the university in relation to your selected group. List these five variables below and note whether they are nominal, ordinal, disrete, or continuous variables. For example:

- ex: Average Faculty Salary (AVGFACSAL) / Continuous
- ex: Control (CONTROL) / Nominal
- ex: Percentage of Degrees Awarded in Social Sciences (PCIP45) / Continuous
- ex: Number of graduate students (GRADS) / Discrete
- ex: Locale (LOCALE2) / Nominal

LIST YOUR FIVE VARIABLES BELOW
- 
-
- 
- 
- 

2. For each of these variables, determine how the values would impact the score of the university according to your metric. 

For Continuous or Discrete variables determine whether higher values will *positively* or *negatively* impact the score of the university in your metric. For example: 
- ex: Average Faculty Salary Score: higher salaries will positively impact the score
- ex: Percentage of Degrees Awarded in Social Sciences: higher percentages will positively impact the score
- ex: Number of graduate students: higher numbers will positively impact the score

For Nominal or Ordinal variables, list the desired value. (Note: For the sake of simplicity, we will only be selecting one desired value from the variable here. Some of you may wish to rank a series of values for a variable. If this is the case, I'm happy to go over how to do this in office hours.)
- ex: Control: Public Universities (1) 
- ex: Locale: Mid-size City (2)

LIST YOUR FIVE DETERMINATIONS BELOW
- 
- 
- 
- 
- 

3. Calculate the top ten universities according to your metric. I will provide most of the code for this. Here is an example:
```{r}
scorecard_ranked <- scorecard %>%
  mutate(ranked = ((rescale(PCIP45)*2) + rescale(AVGFACSAL) + rescale(GRADS)))

scorecard_ranked %>% 
  filter(CONTROL == 1 & LOCALE == 12) %>% 
  select(INSTNM, AVGFACSAL, PCIP45, GRADS, ranked) %>% 
  top_n(10, ranked) %>% 
  arrange(desc(ranked))
```

ENTER YOUR VARIABLES AND VALUES BELOW
```{r}
scorecard_ranked <- scorecard %>%
  mutate(ranked = (rescale(_____) + rescale(_____) - rescale(_____)) #Fill in your discrete and continuous variables here. First add the rescaled variables in which higher values will positively impact the score (you may need to add more) and then subtract the rescaled variables in which higher values will negatively impact the score (if you don't have any, then remove the subtraction part above.) This will create a new column in the dataset that ranks each university according to the numeric values in your metric. Note that you may weight variables differently (as I have with PCIP45 in the example above by multiplying by a weight; this is not required)

scorecard_ranked %>% 
  filter(_____ == _____ & _____ == _____) %>% #Fill in your ordinal variables here and the desired values. (Depending on how many nominal variables you chose, you may need to remove the second condition or add more conditions)
  select(INSTNM, _____, _____, _____, _____, _____, ranked) %>% #Fill in your five values
  top_n(10, ranked) %>% #This will extract universities with the highest scores in ranked
  arrange(desc(ranked))
```


**IMPORTANT NOTE**: If you get the following error:
no non-missing arguments to min; returning Infno non-missing arguments to max; returning -Infno non-missing arguments to max; returning -Inf[1] <0 rows> (or 0-length row.names)

It means that you have selected a variable in which data was not collected for that year. This happens in data science. We rely on variables for which we may just not have data available. If this happens, please note that variable below, and remove it from your metric above. 
Missing data:
- 


4. List four stakeholders that might be interested in this metric. For each write a setence explaining their interest in the metric, as well as a sentence explaining what motivates this interest.  

- ex: Real estate businesses: Real estate businesses would be interested in a metric for best universities for faculty members under 30 because it could forecast where young faculty members may be moving, helping them discern where they may want to build out or renovate new apartment buildings. This interest is profit-motivated. 

LIST YOUR FOUR STAKEHOLDERS AND EXPLANATIONS BELOW
- 
-
- 
- 

5. If universities were to become fixated on this metric, how might it change how they operate, the programs they offer, or the way they serve students? If universities were to begin evolving in order to improve their scores in relation to this metric, what unintended consequences could it have on university faculty, staff, students, or outside community members? How would it impact the data inputs for your metric? Please respond in 200-300 words, referencing specific variables outlined in your metric. 

ENTER YOUR RESPONSE BELOW

