---
title: "R Homework 2"
output: html_notebook
---


Run the code chunk below to load the tidyverse library.
```{r}
library(tidyverse)
```


Run the code chunk below to create the scorecard dataframe. Don't worry about the the "NAs introduced by coercion" listed below. 
```{r}
#Read in the csv.
scorecard <- read.csv("Scorecard/MERGED2017_18_PP.csv", stringsAsFactors = FALSE)

#Convert all values entered as "NULL" to NAs. 
is.na(scorecard) <- scorecard == "NULL"

#Convert columns 12 - 1899 to numeric.
for(i in c(12:1899)){
  scorecard[,i]<-as.numeric(scorecard[,i])
}
```


1. Reference the data dictionary to determine the unique key of the scorecard dataset. Confirm that this is a unique key by filling in the code below.
```{r}
n_unique_key <- 
  _____ %>%
  _____(_____) %>%
  _____
  
n_row <- _____(_____)
  
n_unique_key == n_row
```
2. Referencing the About section of the Full Documentation, as well as the Accuracy and Privacy section, summarize below what each observation in the dataset represents. (In other words, what does the unique key refer to?)


3. What is the average in-state tuition and fees for all public universities in California? What is the average out of state tuition and fees for all public universities in California?
```{r}
#Calculate average in-state tuition and fees for all public universities in California by filling in the *blanks*
scorecard %>%
  _____(_____ == _____) %>%
  _____(_____ == _____) %>%
  _____(avg_tuition_fees = _____(_____, _____))

#Calculate average out-of-state tuition and fees for all public universities in California by filling in the entire *row*
scorecard %>%
  _____ %>%
  _____ %>%
  _____
```
4. Referencing the Full Documentation, which students are included in the calculation for in-state and out-of-state tuition and fees?


5. What are the names of the Hispanic-serving institutions in California, and what share of their undergraduate enrollment is Hispanic? (Fill in lines) What are the names of the women-only institutions in California, and what share of their undergraduate enrollment is women? (Fill in entire rows)
```{r}
#Display the names of universities designated as Hispanic-serving institutions in CA, alongside the share of their undergrdaute enrollment that is Hispanic
scorecard %>%
  _____(_____ == _____) %>%
  _____(_____ == _____) %>%
  _____(_____, _____)

scorecard %>%
  _____ %>%
  _____ %>%
  _____
```
6. Referencing the Full Documentation, determine the criteria by which an institution can become designated as a Hispanic-serving institution. (Hint: You should look in footnote _____, click on the hyperlink. Scroll to the top of this page, and under the Program heading, click the link referring to Hispanic-Serving institutions. In the blue box at the top of this page, click "Eligibility", and in the text on this page there will be a link to the definition of an HSI.) Write the criteria that makes an institution Hispanic-serving below. 


7. What is the name of the university in California with the lowest admission rate? (Fill in lines)
```{r}
#Calculate the lowest admission rate in California
scorecard %>%
  _____(_____ == _____) %>%
  _____(min_admission_rate = _____(_____, _____))
```
```{r}
#Display the name of the university with the lowest admission rate in California
scorecard %>%
  _____(_____ == _____) %>%
  _____(_____)
```
8. Referencing this sheet: https://nces.ed.gov/pubs2019/2019037.pdf, how is admission rate calculated? What institutions are included and excluded in this calculation?

9. What is the name of the university in California with the highest average faculty salary, and what are its instructional expenditures per full-time student? (Fill in rows.)
```{r}
#Calculate the highest average faculty salary in California
scorecard %>%
  _____ %>% 
  _____
```

```{r}
#Display the name of the university with the highest average faculty salary in California, along with the instructional expenditures per full-time students
scorecard %>%
  _____
  _____
```

10. Referencing the Full Documentation, how was average faculty salary calculated? Who was included and excluded? (see also: https://surveys.nces.ed.gov/ipeds/VisGlossaryPopup.aspx?idlink=996) According to this definition when would an adjunct, lecturer, or post-doc be included in this calculation. 
