---
title: "311 Intro to Plotting in R"
output: html_notebook
---

```{r}
library(tidyverse)
library(lubridate)

sf311_zip <- unzip("311/2018.zip")
sf311 <- read.csv(sf311_zip[1]) 
sf311$Opened <- mdy_hms(sf311$Opened)
sf311$Closed <- mdy_hms(sf311$Closed)
sf311$Updated <- mdy_hms(sf311$Updated)
sf311$Category <- tolower(sf311$Category)
sf311$Police.District <- tolower(sf311$Police.District)
sf311$Category <- gsub("general request - ", "", sf311$Category)
sf311 <- sf311 %>%
  mutate(Month = as.character(month(Opened)))
```


To produce plots in R, we will use a package in the Tidyverse called ggplot(). To create any type of plot in ggplot, you should follow this formula:

For plots with only an x-axis:
data %>% ggplot(aes(x=VARIABLE)) + [PLOT_TYPE] 

For plots with an x and y-axis: 
data %>% ggplot(aes(x=[VARIABLE], y=[VARIABLE])) + [PLOT_TYPE] 

ggplot produces plots according to a "grammar of graphics." Just like how the Tidyverse allows us to string together functions with the pipe, ggplot allows us to string together plot attributes with a plus ('+') sign. Using the plus sign, we can do things like:
- add titles 
data %>% ggplot(aes(x=[VARIABLE], y=[VARIABLE])) + [PLOT_TYPE] + labs(title = "[TITLE]")

- add axis labels
data %>% ggplot(aes(x=[VARIABLE], y=[VARIABLE])) + [PLOT_TYPE] + labs(x = "[XAXIS]", y = "[YAXIS]")

- set x and y axis limits
data %>% ggplot(aes(x=[VARIABLE], y=[VARIABLE])) + [PLOT_TYPE] + xlim(0,1000) + ylim(0,200)

- display tick labels vertically rather than horizontally
data %>% ggplot(aes(x=[VARIABLE], y=[VARIABLE])) + [PLOT_TYPE] + theme(axis.text.x = element_text(angle = 90, hjust = 1))

Today, we are going to work with two plot types:
1. Column Charts
2. Line Charts

These charts are useful when you want to plot categorical variables on one axis and numeric variables on the other axis. What categorical variables do we have in 311? What numeric variables do we have in 311?

To produce our plots, we are going to count the number of times certain values appear within variables in the dataset. To do this, we want to use the function _______. 

Calcuate the number of 311 complaints received in each complaint Category. Plot this as a column chart. Add a title.
```{r}
sf311 %>%
  _____(_____) %>%
  _____(Count = _____ %>%
  _____ %>%
  _____(aes(x=_____, y=_____)) + _____ + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Filter the data to the top complaint category above. Calculate the number of 311 complaints received for each request type in this category. Plot this as a column chart. Add a title. 
```{r}
sf311 %>%
  _____(_____ == _____) %>%
  _____(_____) %>%
  _____(Count = _____) %>%
  _____ %>%
  _____(aes(x=_____, y=____)) + ______ + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Calculate the number of 311 complaints received in each Neighborhood in SF. Plot this as a column chart. 
```{r}
sf311 %>%
  _____ %>%
  _____ %>%
  _____ %>%
  _____
```
We can add additional categorical variables to column charts, using the fill aesthetic. When we set the fill aesthetic to another categorical variable, each column in the chart will be broken into different color segments, representing a value within the categorical variable. The size of the colored segement will depend on the number of times that value appears alongside the column value in the dataset. In this way, fill produces *stacked bar charts* that demonstrate the proportion of values in one variable that are present alongside values in a second variable. 


Calculate the number of 311 complaints received by each source for each category. Create a stacked column chart with category columns filled according to the source. 
```{r}
sf311 %>%
  _____(_____, _____) %>%
  _____(Count = _____ %>%
  _____ %>%
  ____(_____(x=____, y=_____, fill=_____)) + _____ + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


Calculate the number of 311 complaints received by each source in each neighborhood. Create a stacked column chart with neighborhood columns filled according to the source.
```{r}
sf311 %>%
  _____ %>%
  _____ %>%
  _____ %>%
  _____
```
We can also create separate plots for each value represented in a particular variable, using the *facet* property. By faceting the plots, we create separate panels of a plot for value present in a variable. 

Plot the number of 311 complaints in each police disrict in San Francisco, and facet the plot according to the complaint category. 
```{r}
sf311 %>%
  _____(_____, _____) %>%
  _____(Count = _____) %>%
  _____ %>%
  _____(_____(_____, _____)) + _____ + _____(~_____) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Calclulate the number of complaints per month in 311. Plot this as a line chart. 
```{r}
sf311 %>%
  _____(_____) %>%
  _____(Count = _____) %>%
  _____ %>%
  _____(_____(_____, _____)) + _____(group=1) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Calclulate the number of complaints per month for each source in 311. Plot this as a stacked line chart. Hint: Instead of fill to fill the columns with a color, we will use the aesthetic color= to designate the color of each line. We also have to ggplot how it should group our variables for each line, which we can do by calling, group= as an aesthetic. 

```{r}
sf311 %>%
  _____(_____, _____) %>%
  _____(Count = _____) %>%
  _____ %>%
  _____(_____(_____, _____, color=_____, group=_____)) + _____ + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Calclulate the number of complaints per month for each category in 311. Plot faceted line charts of the complaints per month for each category. 
```{r}
sf311 %>%
  _____(_____, _____) %>%
  _____(Count = _____) %>%
  _____ %>%
  _____(_____(_____, _____, group=1)) + _____ + _____(~_____) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```
