---
title: "R Notebook"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Variables

Variables are used to store data in R. We can change the values in a variable as needed. We use “<-” to assign a variable.

```{r}
class_name <- "FILL"
class_dept <- "FILL"
class_number <- "FILL"
class_size <- 999

sts_majors <- 999
stats_majors <- 999
other<- 999 

```

Every object in R has a particular *class*, which designates the variable's "type" and how functions can be applied to it. We can check the class of a variable by calling it in "class()".
```{r}
class(class_name)
class(class_size)
class(class_number)
```

If someone drops a class, we can easily change the value of the variable.
```{r}
class_size <- class_size - 1
```

We can also perform calculations on variables, such as addition, subtraction, multiplication, and division. Checking whether variables are greater than, less than, or equal to each other will return TRUE or FALSE.
```{r}

sts_stats_majors <- sts_majors + stats_majors

residual <- class_size - sts_stats_majors

sts_majors > stats_majors
residual == other
```

How would I use these variables to determine the percentage of the class that are STS majors? Store it in a variable.
```{r}

```

We can also perform operations on strings - concatenating them with the *paste* function. For instance, we can paste the class_dept string with the class_number string to create a class code. When we do this, we need to tell R what characters should separate the strings.

```{r}
class_code <- paste(class_dept, class_number, sep=" ")
class_code
```


If I wanted to change the class of class number from a character to a numeric, I would call "as.numeric()". Why might I want to do this?
```{r}
class_number <- as.numeric(class_number)
class(class_number)
```

## Vectors

A vector is a set of values (organized like a list) that are all of the same type. A vector can be of type integer, double, characters, logical, for example. We create a vector by placing a set of values in "c(___)". Here c stands for "combine" - indicating that we are combining values.  
```{r}
birth_months <- c(4, 7, 12, 3, 1)
first_letter_name <- c("A","B","C","D","E")
time_on_phone <- c(34, 90, 2, 6, NA)
```

We can measure the length of a vector by calling the function *length*. This counts how many values are listed in the vector. 
```{r}
length(birth_months)
```

We can also extract values from specific positions in the vector by referencing the index in brackets. So let's say I want to extract the third value in the vector. 
```{r}
birth_months[3]
```

To extract all values except the value in a particular position, we will use the "-" sign before the index in brackets. So let's say I want to extract all values except the third in the vector.
```{r}
birth_months[-3]
```

We can also extract a range of values from specific positions in the vector by referencing that range of indexes in brackets separated by a ":". So let's say I want to extract the first through the third value in the vector.
```{r}
first_letter_name[1:3]
```

Finally, to extract values from a non-sequential combination of specific positions in the vector, we can reference each of their indexes in brackets in "c()".  
```{r}
first_letter_name[c(1,3,5)]
```

We can perform operations on vectors - finding their max, their min, their sum, their average, for example. However, we may get an error if we have any empty values in the dataset. To avoid this, we need to communicate to R to remove NA values. 
```{r}
max(time_on_phone)
min(time_on_phone)
sum(time_on_phone)
mean(time_on_phone)

max(time_on_phone, na.rm=TRUE)
min(time_on_phone, na.rm=TRUE)
sum(time_on_phone, na.rm=TRUE)
mean(time_on_phone, na.rm=TRUE)
```

We can also get a list of all of the unique values in a vector by calling "unique()". So, for instance, if we had a vector of values 1, 4, 5, 1, 4, the unique values would be 1, 4, and 5. 
```{r}
unique(first_letter_name)
```


## Matrices

We won't be using matrices much in this course. However, knowing this data type will help you understand its differences from the data types we will use. Let's create a second numeric vector. 
```{r}
birth_months2<-c(999, 999, 999, 999, 999)
```

By binding this together with birth_months, we create a matrix - or a 2-dimensional collection of elements of all the same type.
```{r}
birth_months_matrix<-rbind(birth_months, birth_months2)
birth_months_matrix
```

We can determine the number of rows in the matrix by calling *nrow* and we can determine the number of columns in the matrix by calling *ncol*.
```{r}
nrow(birth_months_matrix)
ncol(birth_months_matrix)
```

We can also extract values from specific positions in the vector by referencing the row position and the column position of the number in brackets. The format for doing so is [row, column] So let's say that we want to extract the value in the second row, third column.
```{r}
birth_months_matrix[2,3]
```
To extract the entire second row, we would leave the column position blank. This would return a *vector* of values stored in the second row of the matrix.
```{r}
birth_months_matrix[2,]
```
And to extract the entire third column, we would leave the row position blank. This would return a *vector* of values stored in the third column of the matrix. 
```{r}
birth_months_matrix[,3]
```
Write code to get the value(s) in:
* the first row, fourth column?
* the second row, first three columns?
* the first row, second and fourth column?

What type of object in R does each of these return?
```{r}
birth_months_matrix[1,4]
birth_months_matrix[2,1:3]
birth_months_matrix[1,c(2,4)]
```

## Lists

I will only briefly go over *lists* because, for the most part, we will not be using them in this class. *Lists* are collections of objects in R. For instance, you can have a collection of numeric vectors, character matrices, and other lists. You can assign names to the objects in lists so that you can more easily reference them. Once a name has been assigned to the object, you can reference it by listing the name of the list followed by the *$* followed by the name of the object. 

```{r}
first_list<-list(x=first_letter_name, y=time_on_phone, z=birth_months_matrix)

first_list$x[2:3]
first_list$y[c(1,4)]
first_list$z[2,-2]
```

## Data Frames

Most of the datasets that we work with will have columns of different types. There might be a column of names of type character, and another column of ages of type numeric.

Instead we will be working with *data frames*. A data frame is a lot like what you imagine when you think of a spreadsheet. It is a list of *vectors* of all the same length. 

```{r}
scorecard <- read.csv("_____")
```

When working with very large datasets, we need tools to help us get a sense of the dataset without having to load the entire data frame. For instance, we can view the first 6 rows of the dataset by calling "head()".
```{r}
head(scorecard)
```

"dim()" will tell us the dimensions of the data frame - i.e. the number of rows and the number of columns in the data frame. 
```{r}
dim(scorecard)
```

"str()" provides a great deal of information about the observations in the data frame, including the number of variables, the number of observations, the variable names, their data types, and a list of observations.
```{r}
str(scorecard)
```

Just like as we had done with matrices, we can extract particular rows and columns in a data frame by referring to their indexes in brackets. 
```{r}
scorecard[2,4]
```

The scorecard dataset is a great example of a dataset for which we we're not going to want to have to determine the column index number to extract the values from it. With almost 1000 columns, this could be incredibly tedious. Instead, we can refer to the column name using the same "$" notation that we discussed above. To see a list of column names, we could refer to the data dictionary. We could also use the function "colnames()".
```{r}
colnames(scorecard)
```

