---
title: "R Review - Stop, Question, and Frisk"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---
```{r}
library(tidyverse)
sqf_zip <- unzip("SQF/sqf-2011-csv.zip")
sqf_2011 <- read.csv(sqf_zip, stringsAsFactors = FALSE) 
sqf_2011_race_cat <- read.csv("SQF/sqf_race_categories.csv", stringsAsFactors = FALSE) 
```

```{r}
#for(i in c(13:15, 17, 21:73, 76:78)){
#  sqf_2011[,i]<-ifelse(sqf_2011[,i]=="Y", 1, 0)
#}

sqf_2011 <- sqf_2011 %>% left_join(sqf_2011_race_cat, by = "race")
```

Added variables for 1) weapon found, 2) arrestsumm, 3) forceused
```{r}
sqf_2011 <- 
  sqf_2011 %>%
  mutate(wpnfound=ifelse((pistol=="Y"|riflshot=="Y"|asltweap=="Y"|knifcuti=="Y"|machgun=="Y"|othrweap=="Y"),"Y","N"))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(arrestsumm=ifelse((sumissue=="Y"|arstmade=="Y"),"Y","N"))

sqf_2011 <- 
  sqf_2011 %>%
  mutate(forceused=ifelse((pf_hands=="Y"|pf_wall=="Y"|pf_grnd=="Y"|pf_drwep=="Y"|pf_ptwep=="Y"|pf_baton=="Y"|pf_hcuff=="Y"|pf_pepsp=="Y"|pf_other=="Y"),"Y","N"))

```

----Added----

```{r}
sqf_2011 %>%
  ggplot(aes(x=age)) +
  geom_histogram(binwidth = 10, boundary = 14, fill = "#7cae00") + 
  xlim(100, 1000) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "NYPD Stops by Age 100+ in 2011", x ="Age (in brackets of 10)", y = "Number of Stops") +
  theme_minimal()

sqf_2011 %>%
  ggplot(aes(x=age)) +
  geom_histogram(binwidth = 10, boundary = 14, fill = "#7cae00") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "NYPD Stops by Age in 2011", x ="Age (in brackets of 10)", y = "Number of Stops") +
  theme_minimal()
  

sqf_2011 %>%
  ggplot(aes(x=age)) +
  geom_histogram(binwidth = 10) 

sqf_2011 %>%
  filter(age > 100 & age < 999) %>% summary()


sqf_2011 %>%
  filter(arstmade == 0 & sumissue == 0) %>% nrow()
```


```{r}
library(lubridate)
sqf_2011$dob1 <- mdy(sqf_2011$dob)
sqf_2011$datestop1 <- mdy(sqf_2011$datestop)

sqf_2011 <- sqf_2011 %>%
  mutate(age1 = (datestop1 - dob1) %/% as.duration(years(1)))

sqf_2011 %>%
  filter(age >= 14 & age <= 24) #off NYCLU by 4645

sqf_2011 %>%
  filter(age >= 14 & age <= 24) %>%
  filter(age1 < 14 | age1 > 24) %>%
  filter(age1 != 110 & age1 != 0 & age1 < 100)

sqf_2011 %>%
  filter(age1 >= 14 & age1 <= 24) #Not this because excludes all dates coded as december 31, 1900

sqf_2011 %>%
  filter(age1 !=110 & age!=age1) %>% select(age, age1, dob, dob1)
```
```{r height = 10}
sqf_2011 %>%
  ggplot(aes(x = str_to_title(race_cat))) + 
  geom_bar(fill = "#7cae00") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +#Changes x-axis tick labels 90 degrees 
  scale_y_continuous(labels = scales::comma) +
  labs(title = "NYPD Stops by Race in 2011", x = "Race", y = "Number of Stops")
```

```{r}
sqf_2011 %>%
  ggplot(aes(x = str_to_title(race_cat), fill = as.factor(frisked))) + 
  geom_bar(position = "fill") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +#Changes x-axis tick labels 90 degrees 
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent Stops Resulting in Frisk by Race in 2011", x = "Race", y = "Percent Stops", fill = "Frisked?")

sqf_2011 %>%
  ggplot(aes(x = str_to_title(race_cat), fill = as.factor(frisked))) + 
  geom_bar(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +#Changes x-axis tick labels 90 degrees 
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Stops Resulting in Frisk by Race in 2011", x = "Race", y = "Stops", fill = "Frisked?")

```

```{r}
sqf_2011 %>%
  filter(frisked == "Y") %>%
  ggplot(aes(x = str_to_title(race_cat), fill = wpnfound)) + 
  geom_bar(position = "fill") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent Frisks Resulting in Weapon Found by Race in 2011", x = "Race", y = "Percent Frisks", fill = "Weapon Found?")

sqf_2011 %>%
  filter(frisked == "Y") %>%
  ggplot(aes(x = str_to_title(race_cat), fill = wpnfound)) + 
  geom_bar(position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust=1)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Frisks Resulting in Weapon Found by Race in 2011", x = "Race", y = "Frisks", fill = "Weapon Found?")
```


----Added----

How many stops occurred in 2011?
```{r}
sqf_2011 %>%
  _____
```

What was the age of the oldest individual stopped? What was the age of the youngest individual stopped?
```{r}
sqf_2011 %>%
  _____(max_age = _____(_____))

sqf_2011 %>%
  _____

```

In 2011, for how many stops was the officer uniformed? In what percentage of the total stops was the officer uniformed?
```{r}
sqf_2011 %>%
  _____(uniformed = _____(_____))

sqf_2011 %>%
  _____(uniformed = _____(_____)/_____) * 100
```

In 2011, how many stops involved a frisk? What percentage of the total stops involved a frisk?
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____
```

In 2011, how many stops in which an individual was frisked resulted in a weapon found? What percentage of the frisk stops resulted in a weapon found?
```{r}
sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(weapon = _____(_____))

sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(weapon_pct = _____(_____)/_____) * 100
```

In 2011, how many stops resulted in an arrest or summons? What percentage of the total stops resulted in an arrest or summons?
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____
```

In 2011, how many stopped invdividuals refused to present identification?
```{r}
sqf_2011 %>%
  _____(______ == _____) %>%
  _____(refused = _____)
```

How many individuals of each race were stopped in 2011?
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(stopsrace = _____) 
```

How many individuals of each sex were stopped in 2011?
```{r}
sqf_2011 %>%
  _____
```

How many individuals of each sex were stopped in instances where force was used?
```{r}
sqf_2011 %>%
  _____(_____ == _____) %>%
  _____(_____) %>%
  _____(_____ = _____)
```

For each racial group represented in the dataset, what were the 1) total number of frisks, 2) total numbers of stops, and 3) percentage of stops that resulted in a frisk.

For each racial group represented in the dataset, what were the 1) total number of instances a weapon was found, 2) total numbers of stops, and 3) percentage of stops that resulted in a weapon found.
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(frisked=_____(_____), stopped = _____, pctfrisked = _____/_____*100)

sqf_2011 %>%
  _____
```

Which police precinct had the greatest number of stops? Was this the same precinct with the greatest number of frisks?
```{r}
sqf_2011 %>%
  _____(_____) %>%
  _____(stopspct = _____) %>%
  _____ %>%
  _____(_____ == _____(_____))

sqf_2011 %>%
  _____


```


For each precinct represented in the dataset, what were the 1) total number of frisks, 2) total numbers of stops, and 3) percentage of stops that resulted in a frisk.

For each precinct represented in the dataset, what were the 1) total number of instances a weapon was found, 2) total numbers of stops, and 3) percentage of stops that resulted in a weapon found.
```{r}
sqf_2011 %>%
  _____

sqf_2011 %>%
  _____

```



